---
title: "Activity Report"
author: "Leigh E Meeker"
date: "August 7, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


### Load data into R and preprocess


dat is the original dataset  
Update date variable to be date instead of chr  

datnona is the dataset without NA values  

```{r}
library(ggplot2)
dat <- dat <- read.csv("activity.csv", stringsAsFactors = FALSE)
dat$date <- as.Date(dat$date, format = "%Y-%m-%d")

datnona <- dat[!is.na(dat$steps),]
```

### Mean total number of steps taken per day

Calculate the total number of steps taken per day  
```{r}
ttlstepsperday <- aggregate(formula = datnona$steps ~ datnona$date, data = dat, FUN = sum)
names(ttlstepsperday) <- c("Date", "Steps")
print(ttlstepsperday)
```


Histogram of total steps per day  

```{r}
ggplot(ttlstepsperday, aes(x = Steps)) + geom_histogram(bins = 20, col = "blue", fill = "#7fff00")
```

Calculate and report the mean and median of the total number of steps taken per day  

```{r}
mn <- mean(ttlstepsperday$Steps)
md <- median(ttlstepsperday$Steps)
```

The mean steps per day is `r mn`  
The median steps per day is `r md`

### Average daily activity pattern

```{r}
ttlstepsperintv <- aggregate (formula = steps ~ interval, data = datnona, FUN = mean)
```

Time series plot of the 5-minute interval and the average number of steps taken,
averaged accross all days  

```{r}
ggplot(data = ttlstepsperintv, aes(x = interval, y = steps)) + geom_line()
```


Calculate the 5 minute interval, on average across all the days, contains the maximum number of steps  
```{r}
maxintv <- ttlstepsperintv[ttlstepsperintv$steps == max(ttlstepsperintv$steps), ]
```

The 5-minute interval with the maximum number of steps is `r maxintv`  


### Inputing missing values

Calculate the total number of missing values in the dataset  
```{r}
numnas <- sum(is.na(dat$steps))
```

The number of NAs is `r numnas`  

The strategy I used to fill in missing values was to use the mean of that interval over all days  

Create new dataset with all missing data filled in  
I did this in 2 steps.  
1. Create column and put the mean of the interval  
2. Create another column and put either the steps or the mean
```{r}
dat$newsteps <- ave(dat$steps, dat$interval, FUN = function(x) mean(x[!is.na(x)]))
dat$nonnasteps <- replace(dat$steps, is.na(dat$steps), dat$newsteps[is.na(dat$steps)])
```

Make histogram of total number steps each day  
```{r}
ttlstperday <- aggregate(formula = nonnasteps ~ date, data = dat, FUN = sum)
ggplot(ttlstperday, aes(x = nonnasteps)) + geom_histogram(bins = 20, col = "blue", fill = "#7fff00")
```

Calculate mean and median total number of steps taken per day  
```{r}
mn2 <- mean(ttlstperday$nonnasteps)
md2 <- median(ttlstperday$nonnasteps)
```

The mean steps per day is `r mn2`  
The median steps per day is `r md2`


The values at the beginning of the exercise were:  
The mean steps per day is `r mn`  
The median steps per day is `r md`

The difference is (after replace NA - before):  
Mean: `r mn2 - mn`
Median: `r md2 - md`


The impact is surprisingly minimal.


### Differences between activity patterns between weekdays and weekends

Create new factor variable in dataset with two levels: "weekday" and "weekend" 
```{r}
dat$day <- sapply(dat$date, FUN = function(x) {if(weekdays(x) == "Saturday" || weekdays(x) == "Sunday") "weekend" else "weekday"})

##avg steps by interval and weekday/weekend
avgsteps <- aggregate(formula = nonnasteps ~ interval + day, data = dat, FUN = mean)
```

Make a panel plot containing a time series plot of the 5-minute interval and the average number of
steps taken, averaged across all weekday days or weekend days  
```{r}
ggplot(data = avgsteps, aes(x = interval, y = nonnasteps)) + geom_line(size = 1.5, color = "blue")+facet_grid(day ~ .)
```

### End of Report